---
title: "Chapter2Code"
author: "Reilly Brennan"
date: "2025-07-15"
output:
  html_document: default
  pdf_document: default
---
##Start: Load Packages
```{r setup, include=FALSE}
##Phylogenetics and Sequence Analysis
library(adephylo)
library(ape)
library(Biostrings)
library(ggtree)
library(msa)
library(muscle)
library(microseq)
library(phangorn)
library(phylobase)
library(phyloraster)
library(phylotools)
library(phytools)
library(pegas)
library(seqinr)
library(tidytree)
library(treeio)

##Spatial Data and Mapping
library(geodata)
library(geosphere)
library(raster)
library(rnaturalearth)
library(sf)
library(sp)
library(spatialEco)
library(spatstat.explore)
library(spatstat.geom)
library(stars)
library(terra)
library(tidyterra)

##Data Manipulation and Visualization
library(anticlust)
library(ash)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(latexpdf)
library(magrittr)
library(reshape2)
library(stringr)
library(tidyverse)
library(ggcorrplot)


## Biodiversity and occurence data
library(CoordinateCleaner)
library(countrycode)
library(gbif.range)
library(readr)
library(rgbif)

## NEW: Spatial Statistics and GAMs (for revised analysis)
library(mgcv)       # Generalized Additive Models
library(spdep)      # Spatial dependence (Moran's I)
library(car)        # VIF for collinearity
library(MASS)       # Negative binomial
library(vegan)      # Species accumulation curves
library(FNN)        # Nearest neighbor for spatial matching
```


##Part 1. Parasite Species Richness###

Load Data:
- Read in csv with records of parasites from literature review 
- Create coordinate object with each record and then buffer with 0.5 degrees around each point
- Sort by unique records to take a look at the list 
```{r}
par<- read.csv("ColiPar.csv", header=TRUE) #has coordinates lat/long for olivacea sequences (all, cytb and other)
  wgscoor<- par
  coordinates(wgscoor) <- ~Longitude + Latitude
  proj4string(wgscoor)<- CRS("+proj=longlat +datum=WGS84")
  raster::shapefile(wgscoor, "ColiPar.shp", overwrite=T)
  
 #->buffered in qGIS and imported below (0.5 degrees)
  
sort(unique(par$Parasite))
```
Import buffered file of parasite points and use phyloraster package to consider species richness
```{r}
coli_par <- terra::vect("./ColiParBuff/ColiParBuff.shp")

r2_colipar <- shp2rast(coli_par, sps.col = "Parasite", ymask = FALSE, background = 0, 
             resolution = 0.5) # convert to raster with phyloraster
layer_colipar<- as.data.frame(r2_colipar, xy= T) #create df with raster data
srcoli<- rast.sr(x=r2_colipar) #calculate species richness using phyloraster function
```
Create S. America for map overlay
```{r}
Chile<- read_sf("./CHL_adm/CHL_adm0.shp")
Argentina <-read_sf("./ARG_adm/ARG_adm0.shp")
SA <- st_union(Chile, Argentina)
class(SA)
```
Match the parasite species richness raster to the extent and resolution of South America
```{r}
test2_par <- terra::crop(srcoli, SA)

ext <- floor(ext(SA))
nr <- rast(ext, res = res(srcoli))
values(nr) <- 0
testnew <- terra::mask(nr,SA)

tn_ras<-rast(testnew)
t2_ras<-rast(test2_par)

res(tn_ras) <- 0.5
res(t2_ras)<-0.5
crs(t2_ras)<-crs(tn_ras)
testnew<- resample(testnew, tn_ras)
test2_par<-resample(test2_par, t2_ras)
plot(test2_par)
plot_par <- mosaic(test2_par, testnew, fun="max")
srSA <- terra::mask(plot_par,SA)
```
ggplot map for parasite richness
```{r}
par_plot_df<- as.data.frame(srSA, xy=TRUE)
par_plot<- subset(par_plot_df, SR!=0)
par_zero<- subset(par_plot_df, SR==0)

xlim<-c(-80, -50)
ylim<-c(-60, -15)

parmap <- ggplot(data = par_plot) +
    geom_raster(aes(x = x, y = y, fill = `SR`)) +
    scale_fill_gradientn(colours = c("#fada5fff", "orange","darkorange","darkorange", "#aa0000", "red4"),
                       breaks = c(1,2,3,4,5,6,7,8,9,10),
                       labels= c("1", "2", "3", "4", "5", "6", "7", "8","9","10"))+
    geom_raster(data=par_zero, aes(x = x, y = y), show.legend = F)+
   ggtitle(expression(paste(bold("Parasite Species Richness of"), bolditalic (" O. longicaudatus")))) +
   theme(legend.position = "right",legend.box.spacing = unit(0, "pt"), panel.background = element_rect(fill = "white"),
        panel.grid = element_line(color = "gray94"), plot.title = element_text(hjust = 0.5, size=10),
        axis.title = element_text(face="bold")
        ) +
    coord_sf(xlim=xlim, ylim=ylim) +
  labs(x="Longitude", y= "Latitude") 
parmap
```
Make data frame of parasite species richness for data analysis
```{r}
dfa<- as.data.frame(srcoli, xy=T)
```




##Part 2. ANDV Phylogeny 

###Phylogenetic Tree Building- This can be skipped as the .tree file is provided. However, a new tree can be built. The ML algorithm may find a tree slightly different than the one provided in the .tree file, but it won't significantly change the results.

Load the sequence file ANDV.fasta and explore the sequences:
```{r}
fa_andv <- readDNAStringSet("ANDV.fasta")
len <- seqlengths(fa_andv)
len<-as.data.frame(len)
min(len$len)
max(len$len)
##function we will use throughout ##
alignment2Fasta <- function(alignment, filename) {
  sink(filename)
  
  n <- length(rownames(alignment))
  for(i in seq(1, n)) {
    cat(paste0('>', rownames(alignment)[i]))
    cat('\n')
    the.sequence <- toString(unmasked(alignment)[[i]])
    cat(the.sequence)
    cat('\n')  
  }
  
  sink(NULL)
}
```
Align the sequences. The msa package works with ClustalW, ClustalOmega, and MUSCLE alignment algorithms. Save the alignment
Trim the alignment and convert it to DNAbin object
```{r, eval=FALSE}
seqs_andv<- readDNAStringSet("ANDV.fasta")#fasta file classssic format with about 180 cytb seqs

align_andv<- msaMuscle(seqs_andv) #align sequences, there are different parameters here that can be changed but i am using the default for clustalw

alignment2Fasta(align_andv, 'alignmentANDV_may7.fasta') #my function

aligned_andv<- readFasta('alignmentANDV_may7.fasta')
trimANDV_align<- msaTrim(aligned_andv, gap.end = 0.1, gap.mid = 0.9)
trimANDV_align <- msa2mat(trimANDV_align)  # for use with ape::as.DNAbin(msa.mat)
trimANDV_align<- ape::as.DNAbin(trimANDV_align)

```

Create phydat, and test models of substitution. Build Phylogenetic Tree. Save tree.
```{r, eval=FALSE}
andv_dat<-as.phyDat(trimANDV_align)

andvmodels<- modelTest(andv_dat)

fit_andv<- pml_bb(andvmodels, control = pml.control(trace = 0), rearrangements="stochastic")

bs_andv <- bootstrap.pml(fit_andv, bs=1000, optNni=TRUE,trees=TRUE,
                   control = pml.control(trace = 0))

plotBS(midpoint(fit_andv$tree), p = .5, type="p", digits=2, main="Ultrafast bootstrap")

plotBS(midpoint(fit_andv$tree), bs, p = 50, type="p", main="Standard bootstrap")

plotBS(midpoint(fit_andv$tree), bs, p = 50, type="p", digits=0, method = "TBE", main="Transfer bootstrap")
treeandv<- fit_andv$tree
plot(fit_andv$tree)
plot(treeandv)

treebs_andv<-plotBS(fit_andv$tree, bs_andv, type= "n")
plot(treebs_andv)
write.tree(treebs_andv, "ANDVTree_BS.tree")
```

###Calculating Phylogenetic Richness in Geographic Space
Get shapefile with points for each data from occurence matrix 
```{r}
andv_points<- read.csv("ColilargoANDVdata.csv", header=TRUE) #has coordinates lat/long for cytb coli
andv_points[is.na(andv_points)] <- 0
wgscoor<-andv_points
coordinates(wgscoor) <- ~ Longitude+ Latitude
proj4string(wgscoor)<- CRS("+proj=longlat +datum=WGS84")
raster::shapefile(wgscoor, "ANDV.shp", overwrite=T)
#buffered in qGIS and imported below (0.5 degrees)
```

Phylo geo map just to take a fun look at where the ANDV seqs are coming from 
```{r}
treeandv<-read.tree("ANDVTree_BS.tree")
tipsandv<- read.csv("andvtips.csv", header=TRUE) # a table to change the tip labels so they will match the shp file 
pos_id <-match(treeandv$tip.label,tipsandv$old ) #element position

treeandv$tip.label <- tipsandv$new[pos_id]

plot(treeandv)


andv_points<- read.csv("ColilargoANDVdata.csv", header=TRUE)

andv_coords<-data.frame(tip= c(andv_points$Accession),
                        lat= c(andv_points$Latitude),
                        lon= c(andv_points$Longitude))
row.names(andv_coords)<- andv_coords$tip
andv_coords$tip<-NULL

map_andv<- phylo.to.map(treeandv,andv_coords, rotate= T, plot=F,
                        direction="rightwards",regions="Chile")

xlim<-c(-80, -50)
ylim<-c(-60, -15)
plot(map_andv,direction="rightwards",pts=FALSE,
    xlim=xlim,ylim=ylim,ftype="i",
    fsize=0.8,map.bg="lightgreen",
    colors="slategrey")
title( main= "ANDV Phylogeny",line=-2 ,cex.main = 2,   font.main= 4)
```

Import the buffered points, calculate phylogenetic diversity of ANDV
```{r warning=F}
shpandv <- terra::vect("ANDVbuff/ANDVbuff.shp") # collection points with buffers 
r2ANDV <- shp2rast(shpandv,sps.col = "Accession", ymask = FALSE, background = 0, resolution =0.5) # turn shp to raster with phyloraster function

treeandv<- read.tree("ANDVTree_BS.tree") #load phylogenetic tree
tipsandv<- read.csv("andvtips.csv", header=TRUE) # a table to change the tip labels so they will match the shp file 

pos_id_andv <-match(treeandv$tip.label,tipsandv$old ) #element position
treeandv$tip.label <- tipsandv$new[pos_id_andv]

dataprep_andv<- phylo.pres(x=r2ANDV, tree=treeandv, pruning= "tree") # take the data points for which their is phylo and shp. data
pdr_andv <- rast.pd(x = dataprep_andv$x, edge.path = dataprep_andv$edge.path, 
               branch.length = dataprep_andv$branch.length) # get phlyo diversity with function from phyloraster
```

Normalize pd values by number of individuals
```{r}
#create df from raster layer 
layer_andv<- as.data.frame(r2ANDV, xy=T)

#create df from PD values
dfandv_coli<- as.data.frame(pdr_andv$PD, xy=T)
colnames(dfandv_coli)[3]<-"PDandv"

#merge and divide PD by # of individuals
seqANDV<- merge(dfandv_coli, layer_andv, by=c("x", "y"))
seqandv_sum<-seqANDV
seqandv_sum$SeqSum <- rowSums(seqandv_sum[,4:33] )
seqANDV$PDandv<- seqandv_sum$PDandv/seqandv_sum$SeqSum


seqANDV[is.na(seqANDV)]<-0
PD_ANDV<- data.frame(seqANDV[1:3])
r_PDandv<- rast(PD_ANDV, type="xyz")
plot(r_PDandv)
```

Phylogenetic diversity of ANDV Map
```{r}
test2 <- terra::crop(r_PDandv$PDandv, SA)
ext <- floor(ext(SA))
nr <- rast(ext, res = res(r_PDandv$PDandv))
values(nr) <- 0
testnew <- terra::mask(nr,SA)

tn_ras<-rast(testnew)
t2_ras<-rast(test2)

res(tn_ras) <- 0.5
res(t2_ras)<-0.5
crs(t2_ras)<-crs(tn_ras)
testnew<- resample(testnew, tn_ras)
test2<-resample(test2, t2_ras)
plot(test2)
plot<- terra::mosaic(test2,testnew, fun="max")

PDplot_andv <- terra::mask(plot,SA)
color<-colorRampPalette(c("darkgray", "#fada5fff", "gold", "orange","darkorange2" ,"#aa0000")) 
plot(PDplot_andv,
     main= "ANDV Phylogenetic Diversity", ext=c(-80, -50, -60, -15), xlab="Longitude", ylab="Latitude",
     col=color(30))
```

Phylogenetic diversity of ANDV Map GGPLOT
```{r}
andv_plot_df<- as.data.frame(PDplot_andv, xy=TRUE)
andv_plot<- subset(andv_plot_df, PDandv!=0)
andv_zero<- subset(andv_plot_df, PDandv==0)

xlim<-c(-80, -50)
ylim<-c(-60, -15)

andvmap<-ggplot(data = andv_plot) +
    geom_raster(aes(x = x, y = y, fill = `PDandv`)) +
    scale_fill_gradientn(colours = c("#fada5fff", "orange","darkorange2","#aa0000", "red4"),
                       breaks = c(0.05, 0.1, 0.15, 0.2,  max(andv_plot$PDandv)),
                       labels= c( "0.05","0.1","0.15","0.2", "0.22"))+
    geom_raster(data=andv_zero, aes(x = x, y = y), show.legend = F)+
    theme(legend.position = "right",legend.box.spacing = unit(0, "pt"), panel.background = element_rect(fill = "white"),
        panel.grid = element_line(color = "gray94"),
        axis.title = element_text(face="bold"),
        plot.title = element_text(hjust = 0.5, size=10)) +
    coord_sf(xlim=xlim, ylim=ylim) +
  labs(x="Longitude", y= "Latitude")+
    ggtitle(expression(paste(bold("Phylogenetic Richness of ANDV"))))
andvmap
```





##Part 3. Colilargo Phylo

###Phylogenetic Tree Building- This can be skipped as the .tree file is provided. However, a new tree can be built. The ML algorithm may find a tree slightly different than the one provided in the .tree file, but it won't significantly change the results.
Align the sequences. The msa package works with ClustalW, ClustalOmega, and MUSCLE alignment algorithms. Here I am using MUSCLE.
```{r, eval=FALSE}
seqs_coli<- readDNAStringSet("CytBColi_4-22.fasta") 
align_coli<- msaMuscle(seqs_coli) #align sequences, there are different parameters here that can be changed but i am using the default for clustalw
alignment2Fasta(align_coli, 'alignment_july.fasta') #my function
aligned<- readFasta('alignment_july.fasta')
trim_align<- msaTrim(aligned, gap.end = 0.1, gap.mid = 0.9)
```

Build Phylogenetic Tree
```{r, eval=FALSE}
#convert data type
trim_align <- msa2mat(trim_align)  # for use with ape::as.DNAbin(msa.mat)
trim_align<- ape::as.DNAbin(trim_align)
colidat<-as.phyDat(trim_align)

#model testing
colimodels<- modelTest(colidat)
#model fit
fit_coli <- pml_bb(colimodels, control = pml.control(trace = 0))
#bootstrap
bs_coli <- bootstrap.pml(fit_coli, bs=1000, optNni=TRUE,trees=TRUE,
                   control = pml.control(trace = 0))

treebs<-plotBS(fit_coli$tree, bs_coli, type= "n",p=0.5)
plot(treebs)
plotBS(midpoint(fit_coli$tree), p = .5, type="p", digits=2, main="Ultrafast bootstrap")

plotBS(midpoint(fit_coli$tree), bs, p = 50, type="p", main="Standard bootstrap")

plotBS(midpoint(fit_coli$tree), bs, p = 50, type="p", digits=0, method = "TBE", main="Transfer bootstrap")

plotBS(midpoint(fit_coli$tree), bs_coli, p = 50, type="p", main="Standard bootstrap")

write.tree(treebs, "ColiTreeJuly_bs.tree")
write.tree(fit_coli$tree,"ColiTreeJuly.tree" )
```

###Calculating Phylogenetic Richness in Geographic Space

Get shapefile with points for each data from occurence matrix 
```{r}
cyt<- read.csv("CytBData.csv", header=TRUE) #has coordinates lat/long for cytb coli
cyt[is.na(cyt)] <- 0
wgscoor<- cyt
coordinates(wgscoor) <- ~Longitude+Latitude
proj4string(wgscoor)<- CRS("+proj=longlat +datum=WGS84")
raster::shapefile(wgscoor, "CytBColi.shp", overwrite=T)
#buffered in qGIS and imported below (0.5 degrees)
```

Colilargo Phyloraster
```{r warning=F}
shpcoli <- terra::vect("CytColiBuff/CytColiBuff.shp") # collection points with buffers 
r2 <- shp2rast(shpcoli,sps.col = "Accession", ymask = FALSE, background = 0, resolution =0.5) # turn shp to raster with phyloraster function
tree<- read.tree("ColiTreeJuly.tree") #load phylogenetic tree

tips<- read.csv("tips_april23.csv", header=T)
 # make tips match 
pos_id <-match(tree$tip.label, tips$old)#element position
colitree<-tree
colitree$tip.label <- tips$new[pos_id] #here sorting by pos_id

plot(colitree)
##prid for phyloraster##
dataprep<- phylo.pres(x=r2, tree=colitree, pruning= "tree") # take the data points for which their is phylo and shp. data

ed <- rast.ed(dataprep$x, colitree)
pdr <- rast.pd(x = dataprep$x, edge.path = dataprep$edge.path, 
               branch.length = dataprep$branch.length) # get phlyo diversity with function from phyloraster
plot(pdr$PD)
```

Normalize pd values by number of individuals
```{r}
#create df from raster layer of colilargo occ.
layer_coli<- as.data.frame(r2, xy= T)

#create dfs from PD values
dfpd_coli<-as.data.frame(pdr$PD, xy=T)

mean(dfpd_coli$PD[dfpd_coli$PD != 0])

#merge and divide PD by # of individuals
seqPD<- merge(dfpd_coli, layer_coli, by=c("x", "y"))
seqPDsum<-seqPD
seqPDsum$SeqSum <- rowSums(seqPDsum[,4:241] )
seqPD$PD<- seqPDsum$PD/seqPDsum$SeqSum

PD_colilargo<-data.frame(seqPD[1:3])
max(PD_colilargo$PD)
min(PD_colilargo$PD[PD_colilargo$PD != 0])

r_PDcolilargo<-  rast(PD_colilargo, type="xyz")
plot(r_PDcolilargo)
```

Make map pretty use shp files of chile and arg
```{r}
Chile<- read_sf("./CHL_adm/CHL_adm0.shp")
Argentina <-read_sf("./ARG_adm/ARG_adm0.shp")
SA <- st_union(Chile, Argentina)
```

Phylogenetic diversity Colilargo Map
```{r}
test2 <- terra::crop(r_PDcolilargo$PD, SA)

ext <- floor(ext(SA))
nr <- rast(ext, res = res(test2))
raster::values(nr) <- 0
testnew <- terra::mask(nr,SA)

tn_ras<-rast(testnew)
t2_ras<-rast(test2)

res(tn_ras) <- 0.5
res(t2_ras)<-0.5
crs(t2_ras)<-crs(tn_ras)
testnew <- resample(testnew, tn_ras)
test2<-resample(test2, t2_ras)
plot(test2)
plot<- mosaic(test2,testnew, fun="max")

PDplot <- terra::mask(plot,SA)

color<-colorRampPalette(c("gray","lightyellow","yellow", "orange", "darkorange2","#aa0000")) 
plot(PDplot,
     main= "Colilargo Phylogenetic Diversity", ext=c(-80, -50, -60, -15), xlab="Longitude", ylab="Latitude",
     col=color(100))
```

Plotting Phylo Div Colilargo GGPLOT
```{r}
pd_plot_df<- as.data.frame(PDplot, xy=TRUE)
pd_plot<- subset(pd_plot_df, PD!=0)
max<-subset(pd_plot_df, PD=0.1147386)
pdzero<- subset(pd_plot_df, PD==0)
xlim<-c(-80, -50)
ylim<-c(-60, -15)


pdmap<-ggplot(data = pd_plot) +
    geom_raster(aes(x = x, y = y, fill = `PD`)) +
    scale_fill_gradientn(colours = c("#fada5fff", "orange","darkorange2","#aa0000", "red4" ),
                       breaks = c( 0.001, 0.01, 0.02, 0.03, 0.04),
                       labels= c( "0.001", "0.01", "0.02", "0.03", "0.04"))+
    geom_raster(data=pdzero, aes(x = x, y = y), show.legend = F)+
    
    theme(legend.position = "right",legend.box.spacing = unit(0, "pt"), panel.background = element_rect(fill = "white"),
        panel.grid = element_line(color = "gray94"),
        axis.title = element_text(face="bold"),
        plot.title = element_text(hjust = 0.5, size=10)) +
    coord_sf(xlim=xlim, ylim=ylim) +
  labs(x="Longitude", y= "Latitude")+
    ggtitle(expression(paste(bold("Phylogenetic Richness of "), bolditalic("O. longicaudatus"))))
pdmap
```









## Part 3.5 Rodent Species Richness
```{r}
rodents <- terra::vect("./shp_poligons/All_poligons.shp")
rodents$specie

r2_rodents <- shp2rast(rodents, sps.col = "specie", ymask = FALSE, background = 0, 
             resolution = 0.5) # convert to raster with phyloraster
layer_rodents<- as.data.frame(r2_rodents, xy= T) #create df with raster data
sr_rodents<- rast.sr(x=r2_rodents) #calculate species richness using phyloraster function
```
Create S. America for map overlay
```{r}
Chile<- read_sf("./CHL_adm/CHL_adm0.shp")
Argentina <-read_sf("./ARG_adm/ARG_adm0.shp")
SA <- st_union(Chile, Argentina)
```
Match the parasite species richness raster to the extent and resolution of South America
```{r}
test2_rodents <- terra::crop(sr_rodents, SA)

ext <- floor(ext(SA))
nr <- rast(ext, res = res(sr_rodents))
values(nr) <- 0
testnew_rodents <- terra::mask(nr,SA)

tn_ras<-rast(testnew_rodents)
t2_ras<-rast(test2_rodents)

res(tn_ras) <- 0.5
res(t2_ras)<-0.5
crs(t2_ras)<-crs(tn_ras)
testnew_rodents<- resample(testnew_rodents, tn_ras)
test2_rodents<-resample(test2_rodents, t2_ras)
plot(test2_rodents)
plot_rodents <- mosaic(test2_rodents, testnew_rodents, fun="max")
rodentsSA <- terra::mask(plot_rodents,SA)
```
ggplot map for parasite richness
```{r}
plot_rodents
rodents_plot_df<- as.data.frame(plot_rodents, xy=TRUE)
rodents_plot<- subset(rodents_plot_df, SR!=0)
rodents_zero<- subset(rodents_plot_df, SR==0)

xlim<-c(-80, -50)
ylim<-c(-60, -15)

rodentsmap <- ggplot(data = rodents_plot) +
    geom_raster(aes(x = x, y = y, fill = `SR`)) +
    scale_fill_gradientn(colours = c("#fada5fff", "orange","darkorange","darkorange", "#aa0000", "red4"),
                       breaks = c(0,5,15,20),
                       labels= c("0", "5", "15", "20"))+
    #geom_raster(data=rodents_zero, aes(x = x, y = y), show.legend = F)+
   ggtitle(expression(paste(bold("Rodent Species Richness of")))) +
   theme(legend.position = "right",legend.box.spacing = unit(0, "pt"), panel.background = element_rect(fill = "white"),
        panel.grid = element_line(color = "gray94"), plot.title = element_text(hjust = 0.5, size=10),
        axis.title = element_text(face="bold")
        ) +
    coord_sf(xlim=xlim, ylim=ylim) +
  labs(x="Longitude", y= "Latitude") 
 rodentsmap
```
Make data frame of parasite species richness for data analysis
```{r}
dfr<- as.data.frame(sr_rodents, xy=T)

```



##Part 4. Combine ANDV, Colilargo, and parasite species richness

Combine PD, SR, latitude
```{r}
layer_coli<- as.data.frame(r2, xy= T)

##with unique coords
dfpd_coli<-as.data.frame(r_PDcolilargo$PD, xy=T)
dfpar_coli<- as.data.frame(srcoli, xy=T)
dfandv_coli<- as.data.frame(r_PDandv$PDandv, xy=T)
dfrodents<- as.data.frame(sr_rodents, xy=T)

colnames(dfandv_coli)[3]<-"PDandv"

dfpar_coli$SR[dfpar_coli$SR==0]<-NA
dfpd_coli$PD[dfpd_coli$PD==0]<-NA
dfandv_coli$PDandv[dfandv_coli$PDandv==0]<-NA

#with matched standard coooooords 
pdcoli<-as.data.frame(PDplot, xy=T)
srcoli<-as.data.frame(srSA, xy=T) 
andvcoli<-as.data.frame(PDplot_andv, xy=T)
rod<-as.data.frame(rodentsSA, xy=T) 

colnames(rod)[3]<-"rod"
colnames(andvcoli)[3]<-"PDandv"
srcoli$SR[srcoli$SR==0]<-NA
pdcoli$PD[pdcoli$PD==0]<-NA
andvcoli$PDandv[andvcoli$PDandv==0]<-NA
rod$rod[rod$rod==0]<-NA


colilargo<- merge(pd_plot, andv_plot, by=c("x","y"))

dfnew<- merge(pdcoli, srcoli,  by= c("x", "y"))
dfnew<- merge(dfnew, andvcoli, by=c("x","y"))
dfnew2<-merge(dfnew, rod, by=c("x","y"))
dfsub<- subset(dfnew2, ! (is.na(PD) & (is.na(SR)) & is.na(PDandv))) ### this is for climate

write.csv(dfsub, "climate_df_rodent.csv")

subdf_coli<- subset(dfnew2, PD!=0 & PDandv!=0) ##NO NA FOR ANDV AND COLI

write.csv(dfnew2, "dfnew2.csv")


```


##Part 5. Nucleotide and Haplotype Diversity of Colilargo

Load Sequence data and cluster all sequences
```{r}
cyt<- read.csv("CytBData.csv", header=TRUE) #info for the sequences
coor<- as.data.frame(cbind(cyt$Longitude, cyt$Latitude))
clusters<- balanced_clustering(coor, 17, method="centroid") #237/17=
cluster<- as.data.frame(clusters)
cyt$cluster<- cluster
```

Align and trim sequences 
```{r }
fa <- readDNAStringSet("CytBColi.fasta")
len <- seqlengths(fa)
len<-as.data.frame(len)
write.csv(len, "lengths.csv")
len<- read.csv("lengths.csv")
sublen<- subset(len, len>800)

subfa<- fa[c(which(names(fa) %in% sublen$X))]

align<-msaMuscle(subfa)
alignment2Fasta(align, 'alignment.fasta') #my function

aligncon<- readFasta('alignment.fasta')

trimmed_align<- msaTrim(aligncon, gap.end = 0.1, gap.mid = 0.9)

msa.mat <-as.data.frame(msa2mat(trimmed_align) )
write.csv(msa.mat, "msa.mat.csv")
msa.mat<-read.csv("msa.mat.csv", header=TRUE)
tips<- read.csv("msalabels.csv", header=TRUE) # a table to change the tip labels
msa.mat$X[msa.mat$X %in% tips$old] <- tips$new
msa.mat$X[msa.mat$X %in% cyt$Accession]<- cyt$cluster$clusters
msa.mat$X<-as.character(msa.mat$X)
```

Caluclate haplotype and nucleotide diversity by cluster
```{r}
#### HAPLOTYPE DIVERSITY BY CLUSTER #####
hap<-list()
  for(i in 1:length(unique(msa.mat$X))){
  name = unique(msa.mat$X)[i]
  df=  filter(msa.mat, X==name)
  mat= as.matrix(df)
  bin = ape::as.DNAbin(mat)
  div= hap.div(bin)
  hap[[i]]<-div
  }

hapdf<-do.call(rbind.data.frame, hap)
hapdf$cluster<- row.names(hapdf)

#### NUCLEOTIDE DIVERSITY BY CLUSTER #####
nuc<-list()
  for(i in 1:length(unique(msa.mat$X))){
  name = unique(msa.mat$X)[i]
  df=  filter(msa.mat, X==name)
  mat= as.matrix(df)
  bin = ape::as.DNAbin(mat)
  div= nuc.div(bin)
  nuc[[i]]<-div
  }

nucdf<-do.call(rbind.data.frame, nuc)
nucdf$cluster<- row.names(nucdf)
```

Now we want to connect the nucleotide and haplotype diversity to values of parasite species richness, ANDV phylogenetic richness, and environmental variables
Need to match lat long from parasites to pd to map back. This finds the lat long in parasites that is closest to pd and then assigns the parasite richness to that coordinate
```{r}
seqPD2<-subset(seqPD, PD!=0) ## seqPD is from line 429 ;)
for (i in 4:ncol(seqPD2)) {
  seqPD2[[i]] <- ifelse(seqPD2[[i]] == 1, seqPD2$PD, 0) ##replace binary vals with PD val
}

pdacc<-(seqPD2[,4:ncol(seqPD2)])
pdacc<-as.data.frame(cbind(unlist(pdacc, use.names = FALSE), rep(names(pdacc), each = nrow(pdacc))))
pdacc<-subset(pdacc, V1 !=0)
write.csv(seqPD2, "SeqPD.csv")

pdacc<- pdacc %>% group_by(V2) %>% top_n(1, V1) %>% distinct(V1, V2, .keep_all = TRUE) #get rid of duplicates
colnames(pdacc)[1] <-"trait"
colnames(pdacc)[2] <-"label"

tree<- read.tree("ColiTreeApril23_bs.tree") #load phylogenetic tree

tips<- read.csv("tips_april23.csv", header=T)
 # make tips match 
pos_id <-match(tree$tip.label, tips$old)#element position
tree$tip.label <- tips$new[pos_id] #here sorting by pos_id

pdtable <- full_join(tree, pdacc, by = 'label')
pd <- as_tibble(pdtable)
pd$trait[is.na(pd$trait)] <- 0
pd$trait<- as.numeric(pd$trait)
pdtree<- as.treedata(pd)

dfpd <- reshape2::melt(pdacc, id = "label")
dfpd$value<- as.numeric(dfpd$value)
```

Parasites
```{r}
seqPD2<- merge(dfpd_coli, layer_coli, by=c("x", "y"))
#match the lat and long from the parasites and pd
new<-dfpar_coli %>%
  rename(Latpar=x, Lonpar=y) %>%
  crossing(seqPD2) %>%
  mutate(dist=sqrt((x-Latpar)^2+(y-Lonpar)^2)) %>%
  group_by(x, y) %>%
  filter(dist==min(dist)) %>%
  ungroup() %>%
  select(x, y, PD, SR)


#merge the dbs 
sr_seq<- merge(new, seqPD2, by=c("x", "y") )
sr_seq<- subset(sr_seq, PD.x !=0 | SR !=0)
for (i in 4:ncol(seqPD2)) {
  seqPD2[[i]] <- ifelse(seqPD2[[i]] == 1, seqPD2$PD, 0) ##replace binary vals with PD val
}
for (i in 6:ncol(sr_seq)) {
 sr_seq[[i]] <- ifelse(sr_seq[[i]] == 1, sr_seq$SR, 0) ##replace binary vals with Sr val
}

spacc<-(sr_seq[,6:ncol(sr_seq)])
spacc<-as.data.frame(cbind(unlist(spacc, use.names = FALSE), rep(names(spacc), each = nrow(spacc))))
spacc<-subset(spacc, V1 !=0)

spacc<- spacc %>% group_by(V2) %>% top_n(1, V1) %>% distinct(V1, V2, .keep_all = TRUE)
colnames(spacc)[1] <-"trait"
colnames(spacc)[2] <-"label"

df <- reshape2::melt(spacc, id = "label")
df$value<- as.numeric(df$value)## use df for parasite species richness

```

ANDV pd
```{r}
seqPD2<- merge(dfpd_coli, layer_coli, by=c("x", "y"))
#match the lat and long from the parasites and pd
new_andv<-dfandv_coli %>%
  rename(LatPD=x, LonPD=y) %>%
  crossing(seqPD2) %>%
  mutate(dist=sqrt((x-LatPD)^2+(y-LonPD)^2)) %>%
  group_by(x, y) %>%
  filter(dist==min(dist)) %>%
  ungroup() %>%
  select(x, y, PDandv)

#merge the dbs 
andv_seq<- merge(new_andv, seqPD2, by=c("x", "y") )
andv_seq<- subset(andv_seq, PD !=0 | PDandv !=0)

for (i in 5:ncol(andv_seq)) {
 andv_seq[[i]] <- ifelse(andv_seq[[i]] == 1, andv_seq$PDandv, 0) ##replace binary vals with Sr val
}

andv_acc<-(andv_seq[,5:ncol(andv_seq)])
andv_acc<-as.data.frame(cbind(unlist(andv_acc, use.names = FALSE), rep(names(andv_acc), each = nrow(andv_acc))))
andv_acc<-subset(andv_acc, V1 !=0)

andv_acc<- andv_acc %>% group_by(V2) %>% top_n(1, V1) %>% distinct(V1, V2, .keep_all = TRUE)
colnames(andv_acc)[1] <-"trait"
colnames(andv_acc)[2] <-"label"

tree<- read.tree("ColiTreeApril23_bs.tree") #load phylogenetic tree

tips<- read.csv("tips_april23.csv", header=T)
 # make tips match 
pos_id <-match(tree$tip.label, tips$old)#element position
colitree<-tree
colitree$tip.label <- tips$new[pos_id] #here sorting by pos_id
tips<-as.data.frame(tree$tip.label)
colnames(tips)[1] <- "label"
#dfed<- merge(edacc, tips, by=c("label"))
df_andv <- reshape2::melt(andv_acc, id = "label")
df_andv$value<- as.numeric(df_andv$value)
## use df for parasite species richness
```

Total Data frame with points traced back to accession coordinates 
```{r}
###df merge all ###
tot<- merge(dfpd, df, by= c("label"), all=T)
total_df<- merge(tot, df_andv, by=c("label"), all=T)

total_df$variable.x<- NULL
total_df$variable.y<-NULL
total_df$variable<-NULL

colnames(total_df)[2]<-"ColiPD"
colnames(total_df)[3]<-"ColiParasite"
colnames(total_df)[4]<-"andvPD"
colnames(cyt)[4] <-"label"

dftotal<- merge(total_df, cyt, by=c("label"), all=T)
write.csv(dftotal, "dftotal.csv") #tracced to acc
```

get clim data for accession df
```{r}
geodata::geodata_path("/Users/rbrennan/Library/Mobile Documents/com~apple~CloudDocs/Documents/VT/Projects/Chapter2/Olivacea") ##path to your geodata
r<-geodata::worldclim_global(res=10, var="bio")
r<- r[[c(1,12)]]
names(r) <- c("Temp","Prec")
coords<- data.frame(x=dftotal$Longitude, y=dftotal$Latitude)
points<-vect(coords, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")

values <- terra::extract(r, points, xy=TRUE) #get the temp and precip for each point of diversity (genetic, andv, parasite)
df_acc_clim <- cbind.data.frame(values, dftotal) # same order
df_acc_clim$X<-NULL
write.csv(df_acc_clim, "df_acc_clim.csv")
```

Climate analysis by cluster
```{r}

#sr_cluster<- merge(cyt, df_acc_clim, by=c('label'))

sum_sr<- df_acc_clim %>% group_by(cluster$clusters) %>% summarize(sumpar = mean(ColiParasite), latitude= mean(Latitude),
                                                                   longitude= mean(Longitude),
                                                                 coliphy = mean(ColiPD), andvphy = mean(andvPD), 
                                                                 temp=mean(Temp), prec=mean(Prec))

colnames(sum_sr)[1]<-"cluster"
sr_hap <- merge(sum_sr, hapdf, by=c("cluster"))
colnames(sr_hap)[9]<-"hap"
coli_total <- merge(sr_hap, nucdf, by=c("cluster"))
colnames(coli_total)[10]<-"nuc"

ggplot(coli_total, aes(x=hap, y=latitude, color=sumpar))  + theme_bw()+
  labs(title="Haplotype Diversity by latitude",
       x="coliphy", y = "andvphy") +
      scale_color_gradient2(low="#0f46d1ff", mid="#fada5fff",high= "red",limits=c(0,10),midpoint= 5, name="Parasite SR") + geom_point(size=5)

ggplot(coli_total, aes(x=nuc, y=latitude, color= sumpar))  + theme_bw()+
  labs(title="Nucleotide Diversity by latitude",
       x="Nucleotide diversity", y = "latitude") +
      scale_color_gradient2(low="#0f46d1ff", mid="#fada5fff",high= "red",limits=c(0,10),midpoint= 5, name="Parasite SR") + geom_point(size=5)
```







##Part 6. Build geographic and environmental space

get GBIF occurences for Oligoryzomys longicaudatus
```{r, eval=FALSE}
species<-"Oligoryzomys longicaudatus"
gbif_taxon_keys <- 
species %>% # use only first 1000 names for testing
name_backbone_checklist()  %>% # match to backbone
filter(!matchType == "NONE") %>% # get matched names
pull(usageKey)

occ_download(
pred_in("taxonKey", gbif_taxon_keys),
format = "SIMPLE_CSV")
```
clean occurrences
```{r}
 data <- occ_download_get('0000269-250121130708018') %>%
    occ_download_import() 
# select columns of interest
dat <- data %>%
  dplyr::select(species, decimalLongitude, 
                decimalLatitude, countryCode, individualCount,
                gbifID, family, taxonRank, coordinateUncertaintyInMeters,
                year, basisOfRecord, institutionCode)


# remove records without coordinates
dat <- dat %>% filter(!is.na(decimalLongitude)) %>% filter(!is.na(decimalLatitude))
dat$countryCode <-  countrycode(dat$countryCode, 
                                origin =  'iso2c',
                                destination = 'iso3c')
#flag problems
dat <- data.frame(dat)
flags <- clean_coordinates(x = dat, 
                           lon = "decimalLongitude", 
                           lat = "decimalLatitude",
                           countries = "countryCode",
                           species = "species",
                           tests = c("capitals", "centroids",
                                     "equal", "zeros", "countries"))

#Exclude problematic records
dat_cl <- dat[flags$.summary,] ## these the good ones 

#The flagged records
dat_fl <- dat[!flags$.summary,]
```
get world clim for occ data from GBIF
```{r}
geodata::geodata_path("/Users/rbrennan/Library/Mobile Documents/com~apple~CloudDocs/Documents/VT/Projects/Chapter2/Olivacea")
r<-geodata::worldclim_global(res=2.5, var="bio")
plot(r)
r<- r[[c(1,12)]]
names(r) <- c("Temp","Prec")

names(dat_cl)[2] <- "x"
names(dat_cl)[3] <- "y"

coords<- data.frame(x=dat_cl$x, y=dat_cl$y)
points<-vect(coords, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")
plot(points)

values <-  terra::extract(r, points, xy=T) #extract clim variables from 'r' for 'points'
occ_bio <- cbind.data.frame(values, dat_cl) # same order
# where x.1 and y.1 are the originial latlong and x and y are the closest from worldclim

write.csv(occ_bio, "occ_bio.csv")
```
Plot gbif occurences and calculate mean temp and precip and difference from mean for all gbif occurences
```{r}
occ_bio<-read.csv("occ_bio.csv", header=T)
wallace<- data.frame(scientific_name=occ_bio$species, longitude= occ_bio$x.1, latitude=occ_bio$y.1)

sapply(wallace, anyNA)

write.csv(wallace, "ollo_wallace.csv") ##occurences for wallace, to be added to lit review records
occ_bio$X<-NULL
occ_bio$ID<-NULL
occ_bio$species<-gsub(" ","_", occ_bio$species )
#rename columns
names(occ_bio)[3]<- "LonW"
names(occ_bio)[4]<- "LatW"
names(occ_bio)[6]<- "Longitude"
names(occ_bio)[7]<- "Latitude"
#check data by plotting
wgscoor<- occ_bio
coordinates(wgscoor) <- ~Longitude+Latitude
proj4string(wgscoor)<- CRS("+proj=longlat +datum=WGS84")
plot(wgscoor)

plot(occ_bio$Temp, occ_bio$Prec)
```

Calculate Center of range using gbif.range
```{r}
# Download
obs_pt = get_gbif(sp_name = "Oligoryzomys longicaudatus",
                  basis = c("OBSERVATION","HUMAN_OBSERVATION","MACHINE_OBSERVATION"))
# Plot species records
countries = vect(ne_countries(type = "countries",returnclass = "sf"))
plot(countries,col = "#bcbddc")
points(obs_pt[,c("decimalLongitude","decimalLatitude")],pch=20,col="#99340470",cex=1.5)
# Download ecoregion and read
eco_terra = read_bioreg(bioreg_name = "eco_terra", save_dir = NULL)
# Range
range_coli = get_range(occ_coord = obs_pt,
                        bioreg = eco_terra,
                        bioreg_name = "ECO_NAME")
plot(countries,col = "#bcbddc")
plot(range_coli,col = "#238b45",add = TRUE,axes = FALSE,legend = FALSE)

colirange <- terra::as.polygons(range_coli) #have to change to calculate centroid
colirange<-sf::st_as_sf(colirange)

cent<- st_centroid(st_geometry(colirange)) #calculate centroid
coords_cent <- st_coordinates(cent)
```

get world clim data for the parasite/genetic occurences
```{r}
geodata::geodata_path("/Users/rbrennan/Library/Mobile Documents/com~apple~CloudDocs/Documents/VTProjects/Chapter2/Olivacea")
r<-geodata::worldclim_global(res=10, var="bio")

r<- r[[c(1,12)]]
names(r) <- c("Temp","Prec")

subdf<- read.csv("climate_df_rodent.csv", header=T) # from 537 this has the Phylo Div, ANDV div, and Parasite species richness for Colilargo associated with                                                  longitude and latitude
subdf[1]<-NULL
coords<- data.frame(x=subdf$x, y=subdf$y)
points<-vect(coords, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")


values <- terra::extract(r, points, xy=TRUE) #get the temp and precip for each point of diversity (genetic, andv, parasite)
occ_bio_par <- cbind.data.frame(values, subdf) # same order
occ_bio_par[1]<-NULL
occ_bio_par[5]<-NULL
```

Calculate the distance to range centroid geo
```{r}
##CENTROID from gbif.range GEOGRAPHIC##
x <- c(coords_cent[, "X"])
y <- c(coords_cent[, "Y"])

centdf<- as.matrix(data.frame(x,y))
occdf<- as.matrix(data.frame(occ_bio_par$x, occ_bio_par$y))

dist2cent<-distm(occdf,centdf, fun = distHaversine)
occ_bio_par$distcent<- dist2cent
```



##Part 7. Suitability

Suitability logistic
```{r}
Chile<- read_sf("./CHL_adm/CHL_adm0.shp")
Argentina <-read_sf("./ARG_adm/ARG_adm0.shp")
SA <- st_union(Chile, Argentina)

##Transform suitability tif from maxent into map and dataframe
suit<- rast(x="maxent_gbiflit_accepted_block/Oligoryzomys_longicaudatus_fc.L_rm.0.5_logistic.tif")
test2 <- terra::crop(suit, SA)
ext <- floor(ext(SA))
nr <- rast(ext, res = res(test2))
raster::values(nr) <- 0
testnew <- terra::mask(nr,SA)

tn_ras<-rast(testnew)
t2_ras<-rast(test2)

res(tn_ras) <- 0.5
res(t2_ras)<-0.5
crs(t2_ras)<-crs(tn_ras)
testnew <- resample(testnew, tn_ras)
test2<-resample(test2, t2_ras)

plot(test2)
plot_suit<- mosaic(test2,testnew, fun="max")

suit_plot <- terra::mask(plot_suit,SA)
plot(suit_plot)
suit_df<- as.data.frame(suit_plot, xy=TRUE)
names(suit_df)[3]<-"suit"
```
Binary map
```{r}
# Load original binary raster
bin <- rast("maxent_gbiflit_accepted_block/Oligoryzomys_longicaudatus_fc.L_rm.0.5_p10.tif")

# Binarize in case of float values
bin[] <- ifelse(bin[] >= 0.5, 1, 0)

# Crop and mask to SA shape
bin_crop <- crop(bin, SA)
bin_mask <- terra::mask(bin_crop, SA)

# MATCH GRID: Use suit_plot as the template (has correct res, extent, crs, alignment)
template <- suit_plot  # already aligned to SA, 0.5 deg

# --- KEY STEP: aggregate BEFORE resample to preserve sparse 1s ---
# Estimate aggregation factor between bin and template resolution
fact_x <- round(res(template)[1] / res(bin_mask)[1])
fact_y <- round(res(template)[2] / res(bin_mask)[2])
fact <- c(fact_x, fact_y)

# Aggregate with max: if any underlying pixel is 1, keep 1
bin_agg <- aggregate(bin_mask, fact = fact, fun = "max", na.rm = TRUE)

# Align to template exactly (grid matching)
bin_res <- resample(bin_agg, template, method = "near")

# Final mask (in case of slight edge artifacts)
bin_plot <- terra::mask(bin_res, SA)
plot(bin_plot)

# Convert to df
bin_df <- as.data.frame(bin_plot, xy = TRUE)
names(bin_df)[3] <- "binary"

```
database merge
```{r}
######
occ_bio_suit<- merge(suit_df, occ_bio_par, by=c("x", "y"))
#occ_bio_suit2<- read.csv("occ_bio_suit.csv", header=T) ##original file for next steps of plotting and analysis. If analysis re-done at future data, values might be slightly different since occurences are accessed directly from GBIF and algorithms may slightly differ.
#occ_bio_suit$X<-NULL
occ_bio_suit2<- merge(occ_bio_suit, bin_df, by=c("x", "y"))

plot(occ_bio_suit$suit, occ_bio_suit$SR)

suit_plot_nozero<- subset(suit_df, suit!=0)
suitzero<- subset(suit_df, suit == 0)


```
Plots
```{r, warning=F}
xlim<-c(-80, -50)
ylim<-c(-60, -15)
suit_map<-ggplot(data = suit_plot_nozero) +
    geom_raster(aes(x = x, y = y, fill = `suit`)) +
    scale_fill_gradientn(colours = c( "#FFFFCC",  "#FFD966", "#FFB84D", 
            "#FF9933", "#FF6600", "#CC3300", "#990000"),
                       breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                       labels= c("0", "0.2", "0.4", "0.6", "0.8", "1"),
            name="suitability")+
    geom_raster(data=suitzero, aes(x = x, y = y), show.legend = F)+
    
    theme(legend.position = "right",legend.box.spacing = unit(0, "pt"), panel.background = element_rect(fill = "white"),
        panel.grid = element_line(color = "gray94"),
        axis.title = element_text(face="bold"),
        plot.title = element_text(hjust = 0.5)) +
    coord_sf(xlim=xlim, ylim=ylim) +
  labs(x="Longitude", y= "Latitude")+
    ggtitle(expression(paste(bold("Suitability of "), bolditalic("O. longicaudatus"))))

envSuit<- ggplot(occ_bio_suit, aes(x=Temp, y=Prec, color=suit))  + theme_bw()+
  labs(title="Suitability in Environmental Space",
       x="Temperature", y = "Precipitation", size=100) +  scale_color_gradient2(mid="#fada5fff", high="red",name="Suitability") + geom_point(size=2)  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"), plot.title = element_text(size=13, face="bold"))

bin_map<-ggplot(data = bin_df) +
    geom_raster(aes(x = x, y = y, fill = `binary`)) +
    scale_fill_gradientn(colours = c( "#FFFFCC", "#990000"),
                       breaks = c(0, 1),
                       labels= c("0", "1"),
            name="suitability")+
    geom_raster(data=suitzero, aes(x = x, y = y), show.legend = F)+
    theme(legend.position = "right",legend.box.spacing = unit(0, "pt"), panel.background = element_rect(fill = "white"),
        panel.grid = element_line(color = "gray94"),
        axis.title = element_text(face="bold"),
        plot.title = element_text(hjust = 0.5)) +
    coord_sf(xlim=xlim, ylim=ylim) +
  labs(x="Longitude", y= "Latitude")+
    ggtitle(expression(paste(bold("Binary Suitability of "), bolditalic("O. longicaudatus"))))
suit_map
envSuit
```
plots from wallace data direct
```{r}
wallace_df<- as.data.frame (suit, xy=TRUE)
names(wallace_df)[3]<-"suit"
geodata::geodata_path("/Users/rbrennan/Documents/BePhyNE")
r<-geodata::worldclim_global(res=2.5, var="bio")
r<- r[[c(1,12)]]
names(r) <- c("Temp","Prec")

coords<- data.frame(x=wallace_df$x, y=wallace_df$y)
points<-vect(coords, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")


values <- terra::extract(r, points, xy=TRUE) #extract clim variables from 'r' for 'points'
wallace_values <- cbind.data.frame(values, wallace_df) 
wallace_values[6]<-NULL
wallace_values[6]<-NULL
wallace_env<- ggplot(wallace_values, aes(x=Temp, y=Prec, color=suit))  + theme_bw()+
  labs(title="Habitat Suitability in Environmental Space", x="Temperature", y = "Precipitation", size=100) +  scale_color_gradient2(mid="#fada5fff", high="red",name="Suitability") +
   geom_point(size=2)  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"), plot.title = element_text(size=13, face="bold"))
wallace_env
plot<-ggplot() +
  # Plot wallace_values points with color by suit
  geom_point(data = wallace_values, aes(x = Temp, y = Prec, color = suit), size = 2) +
  # Overlay occ_bio_suit as black "x" points
  geom_point(data = occ_bio_suit, aes(x = Temp, y = Prec), shape = 4, color = "black", size = 2) +
  # Optional: continuous color scale and styling
  scale_color_gradient2(mid="#fada5fff", high="red",name="Suitability")  +
  theme_minimal() +
  labs(x = "Temperature", y = "Precipitation", color = "Suitability")
plot
```
south america
```{r}
library(rmapshaper)
Chile<- read_sf("./CHL_adm/CHL_adm0.shp")
Argentina <-read_sf("./ARG_adm/ARG_adm0.shp")
SA <- st_union(Chile, Argentina)
SA_simple<- rmapshaper::ms_simplify(SA, keep = 0.05)
```
Dist to gbif centroid from thinned data
only gbif data used for first suitability analysis, from Chile/Arg in accepted range
```{r}
gbif_thin<- read.csv("Oligoryzomys_longicaudatus_user_raw_GBIF_ChileArg.csv", header=T)
coords<- data.frame(x=gbif_thin$longitude, y=gbif_thin$latitude)
points_gbif<-vect(coords, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")

points_gbif_sf <- sf::st_as_sf(points_gbif)

# Plot South America with points
mapgbif<-ggplot() +
  geom_sf(data = SA_simple, fill = "gray95", color = "black") +
  geom_sf(data = points_gbif_sf, color = "red", size = 0.5) +
  coord_sf(xlim = c(-85, -50), ylim = c(-60, -15), expand = FALSE) +
  theme_minimal() +
  labs(title = "GBIF Accepted over Chile and Argentina",
       x = "Longitude", y = "Latitude")


values <- terra::extract(r, points_gbif, xy=TRUE) #get the temp and precip for each point of diversity (genetic, andv, parasite)
gbif_thin<- cbind.data.frame(values, gbif_thin) # same order

centroid_gbif<- data.frame(Temp = mean(gbif_thin$Temp, na.rm = TRUE),
                       Prec = mean(gbif_thin$Prec, na.rm = TRUE))

# Combine the centroid and occ_bio_suit into a single matrix
combined_gbif <- rbind(centroid_gbif, occ_bio_suit[, c("Temp", "Prec")])

# Use dist() to calculate Euclidean distances
dists_gbif <- dist(combined_gbif, method = "euclidean")

# Extract the distances from the centroid (row 1) to all other rows
# dist() returns the lower triangle of the distance matrix as a vector,
# so we need to extract only the distances from the first row to the rest.
# The first n-1 elements of the result will be those.
dist_to_gbif <- as.matrix(dists_gbif)[-1, 1]

# Add to the data frame
occ_bio_suit$dist_to_gbif <- dist_to_gbif
```
dist to my points 
lit points only
```{r}
litrev<- read.csv("Genetic_Parasite_Coords.csv", header=T)
litrev$Longitude<-as.numeric(litrev$Longitude)
coords<- data.frame(x=litrev$Longitude, y=litrev$Latitude)
points_lit<-vect(coords, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")

points_lit_sf <- sf::st_as_sf(points_lit)

# Plot South America with points
map_lit<-ggplot() +
  geom_sf(data = SA_simple, fill = "gray95", color = "black") +
  geom_sf(data = points_lit_sf, color = "red", size = 0.5) +
  coord_sf(xlim = c(-85, -50), ylim = c(-60, -15), expand = FALSE) +
  theme_minimal() +
  labs(title = "Points from literature review",
       x = "Longitude", y = "Latitude")



values <- terra::extract(r, points_lit, xy=TRUE) #get the temp and precip for each point of diversity (genetic, andv, parasite)
litrev<- cbind.data.frame(values, litrev) # same order
### dif to my POINTS
centroid_lit<- data.frame(Temp = mean(litrev$Temp, na.rm = TRUE),
                       Prec = mean(litrev$Prec, na.rm = TRUE))

# Combine the centroid and occ_bio_suit into a single matrix
combined_lit<- rbind(centroid_lit, occ_bio_suit[, c("Temp", "Prec")])

# Use dist() to calculate Euclidean distances
dists_lit <- dist(combined_lit, method = "euclidean")

# Extract the distances from the centroid (row 1) to all other rows
# dist() returns the lower triangle of the distance matrix as a vector,
# so we need to extract only the distances from the first row to the rest.
# The first n-1 elements of the result will be those.
dist_to_lit <- as.matrix(dists_lit)[-1, 1]

# Add to the data frame
occ_bio_suit$dist_to_lit <- dist_to_lit

```
dist to realized accepted lit and gbif
```{r}
allacc<- read.csv("Oligoryzomys_longicaudatus_processed_occs.csv", header=T)

coords_allacc<- data.frame(x=allacc$longitude, y=allacc$latitude)
points_allacc<-vect(coords_allacc, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")

points_allacc_sf <- sf::st_as_sf(points_allacc)

# Plot South America with points
map_allacc<-ggplot() +
  geom_sf(data = SA_simple, fill = "gray65", color="gray65") +
  geom_sf(data = points_allacc_sf, color = "red", size = 0.5) +
  coord_sf(xlim = c(-85, -50), ylim = c(-60, -15), expand = FALSE) +
  theme_minimal() +
  labs(title = "Points from GBIF and Lit in accepted range",
       x = "Longitude", y = "Latitude")

library(ggplot2)
map_allacc <- ggplot() +
  geom_sf(data = SA_simple, fill = "gray65", color = "gray65") +
  geom_sf(data = points_allacc_sf, color = "red", size = 0.5) +
  coord_sf(xlim = c(-85, -50), ylim = c(-60, -15), expand = FALSE) +
  theme_minimal() +
  labs(
    x = "Longitude", 
    y = "Latitude"
  ) +
  theme(
    axis.title = element_text(face = "bold")
  ) +
  scale_x_continuous(
    breaks = seq(-85, -50, by = 5),
    labels = function(x) as.character(x)
  ) +
  scale_y_continuous(
    breaks = seq(-60, -10, by = 10),
    labels = function(y) as.character(y)
  )




values <- terra::extract(r, points_allacc, xy=TRUE) #get the temp and precip for each point of diversity (genetic, andv, parasite)
allacc<- cbind.data.frame(values, allacc) # same order
### dif to my POINTS
centroid_allacc<- data.frame(Temp = mean(allacc$Temp, na.rm = TRUE),
                       Prec = mean(allacc$Prec, na.rm = TRUE))

# Combine the centroid and occ_bio_suit into a single matrix
combined_allacc<- rbind(centroid_allacc, occ_bio_suit[, c("Temp", "Prec")])

# Use dist() to calculate Euclidean distances
dists_allacc <- dist(combined_allacc, method = "euclidean")

# Extract the distances from the centroid (row 1) to all other rows
# dist() returns the lower triangle of the distance matrix as a vector,
# so we need to extract only the distances from the first row to the rest.
# The first n-1 elements of the result will be those.
dist_to_allacc <- as.matrix(dists_allacc)[-1, 1]

# Add to the data frame
occ_bio_suit$dist_to_allacc <- dist_to_allacc


```
dist to full realized centroid:
lit points plus all Gbif chile arg
```{r}
full_rec<-read.csv("Oligoryzomys_longicaudatus_processed_occs_GBIF_Lit_ChileArg.csv", header=T)

coords<- data.frame(x=full_rec$longitude, y=full_rec$latitude)
points_full<-vect(coords, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")

points_full_sf <- sf::st_as_sf(points_full)

# Plot South America with points
map_full<-ggplot() +
  geom_sf(data = SA_simple, fill = "gray95", color = "black") +
  geom_sf(data = points_full_sf, color = "red", size = 0.5) +
  coord_sf(xlim = c(-85, -50), ylim = c(-60, -15), expand = FALSE) +
  theme_minimal() +
  labs(title = "GBIF and Lit Review",
       x = "Longitude", y = "Latitude")

values <- terra::extract(r, points_full, xy=TRUE) #get the temp and precip for each point of diversity (genetic, andv, parasite)
full_rec<- cbind.data.frame(values, full_rec) # same order

centroid_realized<- data.frame(Temp = mean(full_rec$Temp, na.rm = TRUE),
                       Prec = mean(full_rec$Prec, na.rm = TRUE))

# Combine the centroid and occ_bio_suit into a single matrix
combined_realized<- rbind(centroid_realized, occ_bio_suit[, c("Temp", "Prec")])

# Use dist() to calculate Euclidean distances
dists_real <- dist(combined_realized, method = "euclidean")

# Extract the distances from the centroid (row 1) to all other rows
# dist() returns the lower triangle of the distance matrix as a vector,
# so we need to extract only the distances from the first row to the rest.
# The first n-1 elements of the result will be those.
dist_to_realized <- as.matrix(dists_real)[-1, 1]

# Add to the data frame
occ_bio_suit$dist_to_realized <- dist_to_realized

```
Binary environmental space and fundamental centroid
```{r}
coords_bin<- data.frame(x=bin_df$x, y=bin_df$y)
points_bin<-vect(coords_bin, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")
plot(points_bin)

values_bin <- terra::extract(r, points_bin, xy=TRUE) #extract clim variables from 'r' for 'points'
bin_values <- cbind.data.frame(values_bin, bin_df) 
bin_values[6]<-NULL
bin_values[6]<-NULL
bin1<- subset(bin_values, binary!=0)

bin_env<- ggplot(bin1, aes(x=Temp, y=Prec, color=binary))  + theme_bw()+
  labs(title="Binary Suitability in Environmental Space", x="Temperature", y = "Precipitation", size=100) + 
   geom_point(size=2, color="red")  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"), plot.title = element_text(size=13, face="bold"))


# Define centroid as a one-row data frame or matrix
centroid <- data.frame(Temp = mean(bin_values$Temp, na.rm = TRUE),
                       Prec = mean(bin_values$Prec, na.rm = TRUE))

# Combine the centroid and occ_bio_suit into a single matrix
combined <- rbind(centroid, occ_bio_suit[, c("Temp", "Prec")])

# Use dist() to calculate Euclidean distances
dists <- dist(combined, method = "euclidean")

# Extract the distances from the centroid (row 1) to all other rows
# dist() returns the lower triangle of the distance matrix as a vector,
# so we need to extract only the distances from the first row to the rest.
# The first n-1 elements of the result will be those.
dist_to_centroid <- as.matrix(dists)[-1, 1]

# Add to the data frame
occ_bio_suit$dist_to_centroid <- dist_to_centroid
```
distance to fundamental from big Suit analysis
```{r}
bin2 <- rast("maxent_gbiflit_all_block/Oligoryzomys_longicaudatus_fc.L_rm.0.5_p10.tif")

# Binarize in case of float values
bin2[] <- ifelse(bin2[] >= 0.5, 1, 0)

# Crop and mask to SA shape
bin_crop2 <- crop(bin2, SA)
bin_mask2 <- terra::mask(bin_crop2, SA)

# MATCH GRID: Use suit_plot as the template (has correct res, extent, crs, alignment)
template <- suit_plot  # already aligned to SA, 0.5 deg

# --- KEY STEP: aggregate BEFORE resample to preserve sparse 1s ---
# Estimate aggregation factor between bin and template resolution
fact_x2<- round(res(template)[1] / res(bin_mask2)[1])
fact_y2 <- round(res(template)[2] / res(bin_mask2)[2])
fact2 <- c(fact_x2, fact_y2)

# Aggregate with max: if any underlying pixel is 1, keep 1
bin_agg2 <- aggregate(bin_mask2, fact = fact, fun = "max", na.rm = TRUE)

# Align to template exactly (grid matching)
bin_res2 <- resample(bin_agg2, template, method = "near")

# Final mask (in case of slight edge artifacts)
bin_plot2 <- terra::mask(bin_res2, SA)
plot(bin_plot2)
# Convert to df
bin_df2 <- as.data.frame(bin_plot2, xy = TRUE)
names(bin_df2)[3] <- "binary"
###
coords_bin2<- data.frame(x=bin_df2$x, y=bin_df2$y)
points_bin2<-vect(coords_bin2, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")
plot(points_bin2)

values_bin2 <- terra::extract(r, points_bin2, xy=TRUE) #extract clim variables from 'r' for 'points'
bin_values2 <- cbind.data.frame(values_bin2, bin_df2) 
bin_values2[6]<-NULL
bin_values2[6]<-NULL
bin2<- subset(bin_values2, binary!=0)

bin_env2<- ggplot(bin2, aes(x=Temp, y=Prec, color=binary))  + theme_bw()+
  labs(title="Binary Suitability in Environmental Space", x="Temperature", y = "Precipitation", size=100) + 
   geom_point(size=2)  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"), plot.title = element_text(size=13, face="bold"))


# Define centroid as a one-row data frame or matrix
centroid2 <- data.frame(Temp = mean(bin_values2$Temp, na.rm = TRUE),
                       Prec = mean(bin_values2$Prec, na.rm = TRUE))

# Combine the centroid and occ_bio_suit into a single matrix
combined2 <- rbind(centroid2, occ_bio_suit[, c("Temp", "Prec")])

# Use dist() to calculate Euclidean distances
dists2 <- dist(combined2, method = "euclidean")

# Extract the distances from the centroid (row 1) to all other rows
# dist() returns the lower triangle of the distance matrix as a vector,
# so we need to extract only the distances from the first row to the rest.
# The first n-1 elements of the result will be those.
dist_to_centroid2 <- as.matrix(dists2)[-1, 1]

# Add to the data frame
occ_bio_suit$dist_to_centroid2 <- dist_to_centroid2

```
those good good plots
```{r}
geodata::geodata_path("/Users/rbrennan/Documents/BePhyNE")
r<-geodata::worldclim_global(res=2.5, var="bio")
r<- r[[c(1,12)]]
names(r) <- c("Temp","Prec")

coords<- data.frame(x=suit_df$x, y=suit_df$y)
points<-vect(coords, geom=c("x","y"), crs ="+proj=longlat +datum=WGS84")
plot(points)

values <- terra::extract(r, points, xy=TRUE) #extract clim variables from 'r' for 'points'
suit_tp <- cbind.data.frame(values, suit_df) 
suit_tp[6]<-NULL
##continuous

suit_tp <- suit_tp[order(suit_tp$suit), ]

library(ggplot2)
library(ggnewscale)

centroid_all <- rbind(
  data.frame(centroid, source = "Existing Fundamental Centroid"),
  data.frame(centroid_allacc, source = "Realized Centroid"))
# Prepare centroid source factor
centroid_all$source <- factor(centroid_all$source,
                              levels = c("Existing Fundamental Centroid",
                                         "Realized Centroid"))

# Split suitability
suit_zero <- subset(suit_tp, suit == 0)
suit_nonzero <- subset(suit_tp, suit > 0)


cont2<-ggplot() +
  # Suitability == 0 (gray), shown with legend
  geom_point(data = suit_zero, aes(x = Temp, y = Prec, color = "Not Suitable"),
             size = 2) +
  scale_color_manual(
    name= NULL,
    values = c("Not Suitable" = "gray")
  ) +

  new_scale_color() +

  # Suitability > 0 (yellow to red gradient)
  geom_point(data = suit_nonzero, aes(x = Temp, y = Prec, color = suit), size = 2) +
  scale_color_gradient2(mid = "#fada5fff", high = "red", name = "Suitability") +

  #new_scale_color() +

  # Literature X points
  geom_point(data = occ_bio_suit, aes(x = Temp, y = Prec, shape = "Literature Record"),
             color = "black", size = 2) +

  # Centroids: shapes + fill color
  geom_point(data = centroid_all, aes(x = Temp, y = Prec, shape = source, fill = source),
              size = 5, stroke = 0.8) +

  # Shape legend (combined for X and centroids)
  scale_shape_manual(
    name = "Record Type",
    values = c("Literature Record" = 4,
               "Existing Fundamental Centroid" = 21,
               "Realized Centroid" = 24)
  ) +

  # Fill color for centroids (no name to avoid duplication)
  scale_fill_manual(
    values = c(
               "Existing Fundamental Centroid" = "black",
               "Realized Centroid" = "green"
      )
  ) +

  guides(
    shape = guide_legend(override.aes = list(
      fill = c( "black", "green", NA),
      color = "black",
      stroke = 1.2,
      size = 3
    ))
  ) +

  theme_minimal() +
  theme(axis.title.x = element_text(size=14, face="bold", colour = "black"),    
    axis.title.y = element_text(size=14, face="bold", colour = "black"))+
  labs(x = "Temperature", y ="Precipitation") +
  ggtitle(expression(paste(bold("Suitability and Centroids of "), bolditalic("O. longicaudatus"))))
  

```





##Part 8. Final analyses and plots 
plots
```{r}

###
SR<-ggplot(occ_bio_suit, aes(x=rod, y=SR, color=PDandv))  + theme_bw()+
  labs(title="Predictors of Parasite Species Richness",
       x="Suitability", y = "Parasite Species Richness") +
       scale_color_gradient2(low="blue",mid="yellow", high="red",name="Distance to Center of Range") + geom_point()
SR
ANDV<- ggplot(occ_bio_suit, aes(x=distcent, y=PDandv, color=Prec ))  + theme_bw()+
  labs(title="Predictors of ANDV Phylogenetic Richness",
       x="Distance to Center of Range", y = "Phylogenetic Richness of ANDV") +
       scale_color_gradient2(low="blue",mid="yellow", high="red",name="Precipiation") + geom_point()   
ANDV
##########
envANDV<- ggplot(occ_bio_par, aes(x=Temp, y=Prec, color=PDandv))  + theme_bw()+
  labs(title="Phylogenetic Richness of ANDV in Environmental Space",
       x="Temperature", y = "Precipitation", size=100) +  scale_color_gradient2(mid="#fada5fff", high="red",name="PR \nANDV") + geom_point(size=2)  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"), plot.title = element_text(size=13, face="bold")) 
envANDV
envColi<- ggplot(occ_bio_suit, aes(x=Temp, y=Prec, color=PD))  + theme_bw()+
  labs(x="Temperature", y = "Precipitation", size=100) + 
   ggtitle(expression(paste(bold("Phylogenetic Richness of "), bolditalic("O. longicaudatus"))))+
scale_color_gradient2(mid="#fada5fff", high="red",name="PR") + geom_point(size=2)  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"), plot.title = element_text(size=13, face="bold"))

all_accepted_env<- ggplot(allacc, aes(x=Temp, y=Prec))  + theme_bw()+
  labs(x="Temperature", y = "Precipitation", size=100)  + geom_point(size=2, color= "red")  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold")) + 
  geom_point(data = centroid_all, aes(x = Temp, y = Prec, shape = source, fill = source),
              size = 5, stroke = 0.8) +scale_shape_manual(name = "Record Type", values = c("Realized Centroid" = 24)) +scale_fill_manual( values = c(NA, "Realized Centroid" = "green") ) 





binary_env<- ggplot(bin1, aes(x=Temp, y=Prec))  + theme_bw()+
  labs(x="Temperature", y = "Precipitation", size=100)  + geom_point(size=2, color= "red")  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold")) + 
  geom_point(data = centroid_all, aes(x = Temp, y = Prec, shape = source, fill = source),size = 5, stroke = 0.8) + scale_shape_manual( name = "Record Type", values = c( "Existing Fundamental Centroid" = 21) ) + scale_fill_manual(
    values = c(NA, "Existing Fundamental Centroid" = "black"))


envSR<- ggplot(occ_bio_par, aes(x=Temp, y=Prec, color=SR))  + theme_bw()+
  labs(title="Parasite Species Richness Environmental Space",
       x="Temperature", y = "Precipitation", size=100) +  scale_color_gradient2(mid="#fada5fff", high="red",name="SR") + geom_point(size=2)  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"), plot.title = element_text(size=13, face="bold"))
envSR

envRod<- ggplot(occ_bio_par, aes(x=Temp, y=Prec, color=rod))  + theme_bw()+
  labs(title="Parasite Species Richness Environmental Space",
       x="Temperature", y = "Precipitation", size=100) +  scale_color_gradient2(mid="#fada5fff", high="red",name="Rodent\nRichness") + geom_point(size=2)  + theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"), plot.title = element_text(size=13, face="bold"))
envRod
```

Correlation tests
```{r}
plot(occ_bio_suit$y, occ_bio_suit$PDandv)
plot(occ_bio_suit$PDandv,occ_bio_suit$suit)

#haplotype
cor.test(coli_total$andvphy,coli_total$hap, method = "spearman", use = "complete.obs")
cor.test(coli_total$sumpar,coli_total$hap, method = "spearman", use = "complete.obs")
cor.test(coli_total$coliphy,coli_total$hap, method = "spearman", use = "complete.obs")

#nucleotide
cor.test(coli_total$andvphy,coli_total$nuc, method = "spearman", use = "complete.obs")
cor.test(coli_total$sumpar,coli_total$nuc, method = "spearman", use = "complete.obs")
cor.test(coli_total$coliphy,coli_total$nuc, method = "spearman", use = "complete.obs")

##check normality
shapiro.test(occ_bio_suit$Temp)
shapiro.test(occ_bio_suit$Prec)
shapiro.test(occ_bio_suit$PD)
shapiro.test(occ_bio_suit$SR)
shapiro.test(occ_bio_suit$PDandv)
shapiro.test(occ_bio_suit$distcent)
shapiro.test(occ_bio_suit$suit)

##cor tests
cor.test(occ_bio_suit$PDandv,occ_bio_suit$PD, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PDandv,occ_bio_suit$suit, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PDandv,occ_bio_suit$Temp, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PDandv,occ_bio_suit$Prec, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PDandv,occ_bio_suit$distcent, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PDandv,occ_bio_suit$dist_to_centroid, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PDandv,occ_bio_suit$dist_to_realized, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PDandv,occ_bio_suit$SR, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$rod,occ_bio_suit$PDandv, method = "spearman", use = "complete.obs")

##sr
cor.test(occ_bio_suit$SR,occ_bio_suit$PD, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$SR,occ_bio_suit$suit, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$SR,occ_bio_suit$Temp, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$SR,occ_bio_suit$Prec, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$SR,occ_bio_suit$distcent, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$SR,occ_bio_suit$dist_to_centroid, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$SR,occ_bio_suit$dist_to_allacc, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$rod,occ_bio_suit$SR, method = "spearman", use = "complete.obs")

## PD
cor.test(occ_bio_suit$PD,occ_bio_suit$PD, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PD,occ_bio_suit$suit, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PD,occ_bio_suit$Temp, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PD,occ_bio_suit$Prec, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PD,occ_bio_suit$distcent, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PD,occ_bio_suit$dist_to_centroid, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$PD,occ_bio_suit$dist_to_realized, method = "spearman", use = "complete.obs")
cor.test(occ_bio_suit$rod,occ_bio_suit$PD, method = "spearman", use = "complete.obs")



#rodent richness




##scaled multivariate##
df_scaled <- as.data.frame(scale(occ_bio_suit[, -c(1:2)] ))

model_sr <- lm(SR ~ PD + suit + Temp+ Prec+ distcent + dist_to_centroid+rod , data = df_scaled)
summary(model_sr)


model_pdandv <- lm(PDandv ~ PD + suit + Temp+ Prec+ distcent +dist_to_centroid , data = df_scaled)
summary(model_pdandv)

model_pd <- lm(PD ~ suit + Temp+ Prec+ distcent +SR +dist_to_centroid, data = df_scaled)
summary(model_pd)

```

correlation plot
```{r}


df_selected <- occ_bio_suit[, c("x", "y", "suit", "Temp", "Prec", "PD", "SR", "PDandv","rod","distcent","dist_to_centroid","dist_to_allacc")]
# Current column names in df_selected:
# "x", "y", "suit", "Temp", "Prec", "PD", "SR", "PDandv","distcent","dist_to_centroid","dist_to_allacc"

# Define the desired order of columns
new_order <- c("SR", "PD", "PDandv", "rod", "distcent", "dist_to_centroid", "dist_to_allacc", "x", "y", "suit", "Temp", "Prec")

# Reorder the dataframe
df_corr <- df_selected[, new_order]

# Rename for axis labels
colnames(df_corr) <- c(
  "Parasite Species Richness", 
  "Phylogenetic Richness", 
  "ANDV Phylogenetic Richness", 
  "Rodent Species Richness",
  "Distance to Geographic Centroid", 
  "Distance to Fundamental Centroid", 
  "Distance to Realized Centroid",
  "Longitude", 
  "Latitude", 
  "Suitability", 
  "Temperature", 
  "Precipitation"
)

# Run this again
# Compute Spearman correlations and p-values
cor_test_spearman <- function(df) {
  n <- ncol(df)
  cor_mat <- matrix(NA, n, n)
  p_mat <- matrix(NA, n, n)
  colnames(cor_mat) <- colnames(p_mat) <- colnames(df)
  rownames(cor_mat) <- rownames(p_mat) <- colnames(df)
  
  for (i in 1:n) {
    for (j in 1:n) {
      test <- cor.test(df[[i]], df[[j]], method = "spearman")
      cor_mat[i, j] <- test$estimate
      p_mat[i, j] <- test$p.value
    }
  }
  
  list(cor = cor_mat, p = p_mat)
}

results <- cor_test_spearman(df_corr)
cor_matrix <- results$cor
p_matrix <- results$p

# Mask non-significant correlations
cor_matrix[p_matrix >= 0.01] <- NA
diag(cor_matrix) <- NA  # Optional: hide 1.0 on diagonal

my_colors <- c("blue", "white", "#B2182B")  # Blue-white-red

corrplot<-ggcorrplot(cor_matrix,
           method = "square",
           type = "lower",
           lab = TRUE,
           lab_size = 2.8,
           tl.cex = 10,
           tl.srt = 45,
           ggtheme = theme_minimal(base_size = 12),
           colors = my_colors,
           outline.col = "white",
           title = "Significant Spearman Correlations") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    panel.grid = element_blank()
  )

corrplot

```

regression plots
```{r}
SR <- ggplot(occ_bio_suit, aes(x =rod, y = SR, color = PDandv)) +
  theme_bw() +

  # Scatter points
  geom_point() +

  # Pearson correlation line + confidence interval
  geom_smooth(method = "lm", se = TRUE, color = "black") +

  # Axis limits based on data max
  scale_x_continuous(limits = c(0, max(occ_bio_suit$rod, na.rm = TRUE))) +
  scale_y_continuous(limits = c(0, max(occ_bio_suit$SR, na.rm = TRUE))) +

  # Color gradient for distance to center
  scale_color_gradient2(
    low = "blue", mid = "yellow", high = "red",
    name = "ANDV\nPhylogenetic\nRichness"
  ) +

  # Axis labels and plot title
  labs(
    title = "Predictors of Parasite and Viral Diversity",
    x = "Host Community Richness",
    y = "Parasite Species Richness"
  ) +

  # Bold axis and title text
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(face = "bold", size = 10)
  )
 SR
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
ANDV<-
  ggplot(occ_bio_suit, aes(x = distcent, y = PDandv, color = Temp)) +
  theme_bw() +

  # Scatter points
  geom_point() +

  # Pearson correlation line + confidence interval
  geom_smooth(method = "lm", se = TRUE, color = "black") +

  # Axis limits based on data max
  scale_x_continuous(limits = c(0, 1250000)) +
  scale_y_continuous(limits = c(0, max(occ_bio_suit$PDandv, na.rm = TRUE))) +

  # Color gradient for distance to center
  scale_color_gradient2(
    low = "blue", mid = "yellow", high = "red",
    name = "Temperature (C)"
  ) +

  # Axis labels and plot title
  labs(
    title = "Predictors of ANDV Phylogenetic Richness",
    x = "Distance to Range Centroid",
    y = "Phylogenetic Richness of ANDV"
  ) +

  # Bold axis and title text
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(face = "bold", size = 10)
  )
ANDV
```



##Part 9. Spatial Autocorrelation Testing and GAM Analysis

This section addresses reviewer concerns about spatial autocorrelation in the OLS models.
We test for spatial autocorrelation using Moran's I, then fit GAMs with spatial smooth terms.
9.1 Prepare Data for Spatial Analysis
```{r }
# SR Model data
model_vars_sr <- c("SR", "PD", "suit", "Temp", "Prec", "distcent", "dist_to_centroid", "rod", "x", "y")
data_sr <- occ_bio_suit[, model_vars_sr]
data_sr <- data_sr[complete.cases(data_sr), ]

# Fix matrix columns (distcent comes from distm() which returns a matrix)
data_sr$distcent <- as.numeric(data_sr$distcent)
data_sr$dist_to_centroid <- as.numeric(data_sr$dist_to_centroid)

# ANDV Model data
model_vars_andv <- c("PDandv", "PD", "suit", "Temp", "Prec", "distcent", "dist_to_centroid", "x", "y")
data_andv <- occ_bio_suit[, model_vars_andv]
data_andv <- data_andv[complete.cases(data_andv), ]
data_andv$distcent <- as.numeric(data_andv$distcent)
data_andv$dist_to_centroid <- as.numeric(data_andv$dist_to_centroid)

```

### 9.2 Test Spatial Autocorrelation in OLS Residuals (Moran's I)
```{r morans_i_test}

# Fit OLS models (same as original analysis)
ols_sr <- lm(SR ~ PD + suit + Temp + Prec + distcent + dist_to_centroid + rod, data = data_sr)
ols_andv <- lm(PDandv ~ PD + suit + Temp + Prec + distcent + dist_to_centroid, data = data_andv)

# Get residuals
resid_sr <- residuals(ols_sr)
resid_andv <- residuals(ols_andv)

#Method 1: K-Nearest Neighbors with Monte Carlo 

# SR Model
coords_sr <- as.matrix(data_sr[, c("x", "y")])
knn_sr <- knearneigh(coords_sr, k = 4)
knn_nb_sr <- knn2nb(knn_sr)
knn_listw_sr <- nb2listw(knn_nb_sr, style = "W")

moran_sr_mc <- moran.mc(resid_sr, knn_listw_sr, nsim = 999)
cat("SR Model (KNN k=4):\n")
cat("  Moran's I =", round(moran_sr_mc$statistic, 4), "\n")
cat("  p-value =", moran_sr_mc$p.value, "\n")
cat("  Significant:", moran_sr_mc$p.value < 0.05, "\n\n")

# ANDV Model
coords_andv <- as.matrix(data_andv[, c("x", "y")])
knn_andv <- knearneigh(coords_andv, k = 4)
knn_nb_andv <- knn2nb(knn_andv)
knn_listw_andv <- nb2listw(knn_nb_andv, style = "W")

moran_andv_mc <- moran.mc(resid_andv, knn_listw_andv, nsim = 999)
cat("ANDV Model (KNN k=4):\n")
cat("  Moran's I =", round(moran_andv_mc$statistic, 4), "\n")
cat("  p-value =", moran_andv_mc$p.value, "\n")
cat("  Significant:", moran_andv_mc$p.value < 0.05, "\n\n")

#Method 2: Inverse Distance Weighting (ape package)
cat("--- Method 2: Inverse Distance Weighting (ape::Moran.I) ---\n\n")

# SR Model
dist_matrix_sr <- as.matrix(dist(coords_sr))
dist_matrix_sr[dist_matrix_sr == 0] <- 1e-10
inv_dist_sr <- 1 / dist_matrix_sr
diag(inv_dist_sr) <- 0

moran_sr_ape <- ape::Moran.I(resid_sr, inv_dist_sr)
cat("SR Model (Inverse Distance):\n")
cat("  Moran's I =", round(moran_sr_ape$observed, 4), "\n")
cat("  p-value =", format(moran_sr_ape$p.value, scientific = TRUE, digits = 3), "\n")
cat("  Significant:", moran_sr_ape$p.value < 0.05, "\n\n")

# ANDV Model
dist_matrix_andv <- as.matrix(dist(coords_andv))
dist_matrix_andv[dist_matrix_andv == 0] <- 1e-10
inv_dist_andv <- 1 / dist_matrix_andv
diag(inv_dist_andv) <- 0

moran_andv_ape <- ape::Moran.I(resid_andv, inv_dist_andv)
cat("ANDV Model (Inverse Distance):\n")
cat("  Moran's I =", round(moran_andv_ape$observed, 4), "\n")
cat("  p-value =", format(moran_andv_ape$p.value, scientific = TRUE, digits = 3), "\n")
cat("  Significant:", moran_andv_ape$p.value < 0.05, "\n\n")


# Save results
moran_results <- data.frame(
  Model = c("SR", "SR", "ANDV", "ANDV"),
  Method = c("KNN (k=4)", "Inverse Distance", "KNN (k=4)", "Inverse Distance"),
  Morans_I = c(moran_sr_mc$statistic, moran_sr_ape$observed, 
               moran_andv_mc$statistic, moran_andv_ape$observed),
  P_value = c(moran_sr_mc$p.value, moran_sr_ape$p.value,
              moran_andv_mc$p.value, moran_andv_ape$p.value),
  Significant = c(moran_sr_mc$p.value < 0.05, moran_sr_ape$p.value < 0.05,
                  moran_andv_mc$p.value < 0.05, moran_andv_ape$p.value < 0.05)
)
print(moran_results)
write.csv(moran_results, "MoransI_OLS_Results.csv", row.names = FALSE)
```

### 9.3 Fit GAMs with Spatial Smooth Terms

We fit GAMs with two different spatial smooth flexibilities:
- k=30: Less flexible smooth that partially addresses spatial autocorrelation 
while retaining environmental signal
- k=50: More flexible smooth that more fully resolves spatial autocorrelation


```{r gam_models_k30}
gam_sr_k30 <- gam(SR ~ s(x, y, k = 30) + PD + suit + Temp + Prec + distcent + 
                    dist_to_centroid + rod,
                  family = nb(),
                  data = data_sr,
                  method = "REML")

summary(gam_sr_k30)

gam_andv_k30 <- gam(PDandv ~ s(x, y, k = 30) + PD + suit + Temp + Prec + 
                      distcent + dist_to_centroid,
                    family = gaussian(),
                    data = data_andv,
                    method = "REML")

print(summary(gam_andv_k30))
# Save k=30 coefficients
sr_coef_k30 <- as.data.frame(summary(gam_sr_k30)$p.table)
sr_coef_k30$Predictor <- rownames(sr_coef_k30)
sr_coef_k30$k <- 30
write.csv(sr_coef_k30, "GAM_SR_Coefficients_k30.csv", row.names = FALSE)

andv_coef_k30 <- as.data.frame(summary(gam_andv_k30)$p.table)
andv_coef_k30$Predictor <- rownames(andv_coef_k30)
andv_coef_k30$k <- 30
write.csv(andv_coef_k30, "GAM_ANDV_Coefficients_k30.csv", row.names = FALSE)
```

```{r gam_models_k50}
gam_sr_k50 <- gam(SR ~ s(x, y, k = 50) + PD + suit + Temp + Prec + distcent 
                  + dist_to_centroid + rod,
                  family = nb(),
                  data = data_sr,
                  method = "REML")
print(summary(gam_sr_k50))

gam_andv_k50 <- gam(PDandv ~ s(x, y, k = 50) + PD + suit + Temp + 
                      Prec + distcent + dist_to_centroid,
                    family = gaussian(),
                    data = data_andv,
                    method = "REML")

print(summary(gam_andv_k50))
# Save k=50 coefficients
sr_coef_k50 <- as.data.frame(summary(gam_sr_k50)$p.table)
sr_coef_k50$Predictor <- rownames(sr_coef_k50)
sr_coef_k50$k <- 50
write.csv(sr_coef_k50, "GAM_SR_Coefficients_k50.csv", row.names = FALSE)

andv_coef_k50 <- as.data.frame(summary(gam_andv_k50)$p.table)
andv_coef_k50$Predictor <- rownames(andv_coef_k50)
andv_coef_k50$k <- 50
write.csv(andv_coef_k50, "GAM_ANDV_Coefficients_k50.csv", row.names = FALSE)

# Set primary models for downstream analysis (using k=30 as primary)
gam_sr <- gam_sr_k30
gam_andv <- gam_andv_k30
```

### 9.3b Compare k=30 vs k=50 Results
```{r compare_k_values}

# Function to extract p-values for comparison
extract_pvals <- function(model, predictors) {
  coef_table <- summary(model)$p.table
  pvals <- coef_table[predictors, "Pr(>|z|)"]
  if (is.null(pvals)) pvals <- coef_table[predictors, "Pr(>|t|)"]
  return(pvals)
}

# SR Model comparison
predictors_sr <- c("PD", "suit", "Temp", "Prec", "distcent", 
                   "dist_to_centroid", "rod")

sr_k_compare <- data.frame(
  Predictor = predictors_sr,
  k30_pval = round(extract_pvals(gam_sr_k30, predictors_sr), 4),
  k50_pval = round(extract_pvals(gam_sr_k50, predictors_sr), 4)
)
sr_k_compare$k30_sig <- ifelse(sr_k_compare$k30_pval < 0.05, "YES", 
                                ifelse(sr_k_compare$k30_pval < 0.1, 
                                       "marginal", "no"))
sr_k_compare$k50_sig <- ifelse(sr_k_compare$k50_pval < 0.05, "YES",
                                ifelse(sr_k_compare$k50_pval < 0.1, 
                                       "marginal", "no"))
sr_k_compare$robust <- sr_k_compare$k30_sig == sr_k_compare$k50_sig

print(sr_k_compare)

# ANDV Model comparison
predictors_andv <- c("PD", "suit", "Temp", "Prec", "distcent", "dist_to_centroid")

andv_k_compare <- data.frame(
  Predictor = predictors_andv,
  k30_pval = extract_pvals(gam_andv_k30, predictors_andv), 4,
  k50_pval = extract_pvals(gam_andv_k50, predictors_andv), 4
)
andv_k_compare$k30_sig <- ifelse(andv_k_compare$k30_pval < 0.05, "YES",
                                  ifelse(andv_k_compare$k30_pval < 0.1, 
                                         "marginal", "no"))
andv_k_compare$k50_sig <- ifelse(andv_k_compare$k50_pval < 0.05, "YES",
                                  ifelse(andv_k_compare$k50_pval < 0.1, 
                                         "marginal", "no"))
andv_k_compare$robust <- andv_k_compare$k30_sig == andv_k_compare$k50_sig

print(andv_k_compare)

# Save comparison
write.csv(sr_k_compare, "SR_k30_vs_k50_Comparison.csv", row.names = FALSE)
write.csv(andv_k_compare, "ANDV_k30_vs_k50_Comparison.csv", row.names = FALSE)

```

### 9.4 Test if GAM Resolved Spatial Autocorrelation
```{r }
# Get GAM residuals for both k values
gam_resid_sr_k30 <- residuals(gam_sr_k30, type = "deviance")
gam_resid_sr_k50 <- residuals(gam_sr_k50, type = "deviance")
gam_resid_andv_k30 <- residuals(gam_andv_k30, type = "deviance")
gam_resid_andv_k50 <- residuals(gam_andv_k50, type = "deviance")

# Test with KNN method

moran_gam_sr_k30 <- moran.mc(gam_resid_sr_k30, knn_listw_sr, nsim = 999)
cat("k=30: Moran's I =", round(moran_gam_sr_k30$statistic, 4), 
    ", p =", moran_gam_sr_k30$p.value,
    ", Resolved:", moran_gam_sr_k30$p.value >= 0.05, "\n")

moran_gam_sr_k50 <- moran.mc(gam_resid_sr_k50, knn_listw_sr, nsim = 999)
cat("k=50: Moran's I =", round(moran_gam_sr_k50$statistic, 4), 
    ", p =", moran_gam_sr_k50$p.value,
    ", Resolved:", moran_gam_sr_k50$p.value >= 0.05, "\n\n")

cat("--- ANDV Model: Post-GAM Moran's I ---\n\n")

moran_gam_andv_k30 <- moran.mc(gam_resid_andv_k30, knn_listw_andv, nsim = 999)
cat("k=30: Moran's I =", round(moran_gam_andv_k30$statistic, 4), 
    ", p =", moran_gam_andv_k30$p.value,
    ", Resolved:", moran_gam_andv_k30$p.value >= 0.05, "\n")

moran_gam_andv_k50 <- moran.mc(gam_resid_andv_k50, knn_listw_andv, nsim = 999)
cat("k=50: Moran's I =", round(moran_gam_andv_k50$statistic, 4), 
    ", p =", moran_gam_andv_k50$p.value,
    ", Resolved:", moran_gam_andv_k50$p.value >= 0.05, "\n\n")

comparison_df <- data.frame(
  Model = c("SR", "SR", "SR", "ANDV", "ANDV", "ANDV"),
  Specification = c("OLS", "GAM k=30", "GAM k=50", "OLS", "GAM k=30", "GAM k=50"),
  Morans_I = c(round(moran_sr_mc$statistic, 3), 
               round(moran_gam_sr_k30$statistic, 3),
               round(moran_gam_sr_k50$statistic, 3),
               round(moran_andv_mc$statistic, 3),
               round(moran_gam_andv_k30$statistic, 3),
               round(moran_gam_andv_k50$statistic, 3)),
  P_value = c(moran_sr_mc$p.value, 
              moran_gam_sr_k30$p.value,
              moran_gam_sr_k50$p.value,
              moran_andv_mc$p.value,
              moran_gam_andv_k30$p.value,
              moran_gam_andv_k50$p.value),
  Autocorr_Resolved = c(moran_sr_mc$p.value >= 0.05,
                        moran_gam_sr_k30$p.value >= 0.05,
                        moran_gam_sr_k50$p.value >= 0.05,
                        moran_andv_mc$p.value >= 0.05,
                        moran_gam_andv_k30$p.value >= 0.05,
                        moran_gam_andv_k50$p.value >= 0.05)
)
print(comparison_df)
write.csv(comparison_df, "Moran_OLS_vs_GAM_All_Comparison.csv", row.names = FALSE)

# Set references for downstream (use k=30 as primary)
moran_gam_sr <- moran_gam_sr_k30
moran_gam_andv <- moran_gam_andv_k30
```

### 9.5 Variance Inflation Factor (VIF) Analysis
```{r vif_analysis}

# VIF for SR model
vif_sr <- car::vif(ols_sr)
print(round(vif_sr, 2))

# VIF for ANDV model
vif_andv <- car::vif(ols_andv)

print(round(vif_andv, 2))

## Summary table
vif_table <- data.frame(
  Predictor = c(names(vif_sr), names(vif_andv)),
  Model = c(rep("SR", length(vif_sr)), rep("ANDV", length(vif_andv))),
  VIF = round(c(vif_sr, vif_andv), 2)
)
vif_table$Status <- ifelse(vif_table$VIF > 10, "SEVERE (>10)",
                           ifelse(vif_table$VIF > 5, "Moderate (5-10)", "OK (<5)"))

print(vif_table)
write.csv(vif_table, "VIF_Results.csv", row.names = FALSE)
# Check correlation between Prec and dist_to_centroid
cor_prec_dist <- cor(data_sr$Prec, data_sr$dist_to_centroid, use = "complete.obs")

```

### 9.6 Species Accumulation and Sampling Effort Analysis
```{r sampling_effort}

# Load original parasite data
par_data <- read.csv("ColiPar.csv", header = TRUE)

# Round to grid cells
par_data$x_cell <- round(par_data$Longitude / 0.5) * 0.5
par_data$y_cell <- round(par_data$Latitude / 0.5) * 0.5
par_data$cell_id <- paste(par_data$x_cell, par_data$y_cell, sep = "_")

# Effort metrics per cell
effort_df <- par_data %>%
  group_by(x_cell, y_cell, cell_id) %>%
  summarise(
    n_records = n(),
    n_studies = n_distinct(Source),
    n_species = n_distinct(Parasite),
    .groups = "drop"
  )

print(summary(effort_df$n_records))

print(summary(effort_df$n_studies))

# Coefficient of variation
cv_records <- sd(effort_df$n_records) / mean(effort_df$n_records) * 100

###Species accumulation curve##
# Create community matrix
comm_matrix <- par_data %>%
  mutate(present = 1) %>%
  select(cell_id, Parasite, present) %>%
  distinct() %>%
  tidyr::pivot_wider(names_from = Parasite, values_from = present, values_fill = 0) %>%
  tibble::column_to_rownames("cell_id")

# Species accumulation
spec_accum <- specaccum(comm_matrix, method = "random", permutations = 100)

# Chao richness estimate
chao <- specpool(comm_matrix)
cat("\nChao richness estimate:", round(chao$chao, 1), "species\n")
cat("Observed richness:", chao$Species, "species\n")
cat("Detection rate:", round(chao$Species / chao$chao * 100, 1), "%\n")

# Plot species accumulation
plot(spec_accum, ci.type = "polygon", ci.col = "lightblue", col = "blue", lwd = 2,
     xlab = "Number of Grid Cells Sampled",
     ylab = "Cumulative Species Richness",
     main = "Species Accumulation Curve")

```

### 9.7 Visualize GAM Spatial Effects
```{r gam_visualization}
# Create prediction grid
x_range <- range(data_sr$x)
y_range <- range(data_sr$y)
pred_grid <- expand.grid(
  x = seq(x_range[1], x_range[2], length.out = 50),
  y = seq(y_range[1], y_range[2], length.out = 50)
)

# Add mean values for other predictors
pred_grid$PD <- mean(data_sr$PD, na.rm = TRUE)
pred_grid$suit <- mean(data_sr$suit, na.rm = TRUE)
pred_grid$Temp <- mean(data_sr$Temp, na.rm = TRUE)
pred_grid$Prec <- mean(data_sr$Prec, na.rm = TRUE)
pred_grid$distcent <- mean(data_sr$distcent, na.rm = TRUE)
pred_grid$dist_to_centroid <- mean(data_sr$dist_to_centroid, na.rm = TRUE)
pred_grid$rod <- mean(data_sr$rod, na.rm = TRUE)

# Predict SR spatial effect
pred_grid$SR_pred <- predict(gam_sr, newdata = pred_grid, type = "response")

# Plot SR spatial effect
p_sr_spatial <- ggplot(pred_grid, aes(x = x, y = y, fill = SR_pred)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted\nSR", option = "plasma") +
  geom_point(data = data_sr, aes(x = x, y = y, fill = NULL), 
             color = "white", size = 1, alpha = 0.5) +
  labs(title = "GAM Spatial Effect: Parasite Species Richness",
       subtitle = "s(x, y) smooth term visualization",
       x = "Longitude", y = "Latitude") +
  theme_bw() +
  coord_fixed()

print(p_sr_spatial)
ggsave("GAM_Spatial_Effect_SR.png", p_sr_spatial, width = 8, height = 10, dpi = 300)

# Plot for ANDV
pred_grid_andv <- expand.grid(
  x = seq(range(data_andv$x)[1], range(data_andv$x)[2], length.out = 50),
  y = seq(range(data_andv$y)[1], range(data_andv$y)[2], length.out = 50)
)
pred_grid_andv$PD <- mean(data_andv$PD, na.rm = TRUE)
pred_grid_andv$suit <- mean(data_andv$suit, na.rm = TRUE)
pred_grid_andv$Temp <- mean(data_andv$Temp, na.rm = TRUE)
pred_grid_andv$Prec <- mean(data_andv$Prec, na.rm = TRUE)
pred_grid_andv$distcent <- mean(data_andv$distcent, na.rm = TRUE)
pred_grid_andv$dist_to_centroid <- mean(data_andv$dist_to_centroid, na.rm = TRUE)

pred_grid_andv$ANDV_pred <- predict(gam_andv, newdata = pred_grid_andv, type = "response")

p_andv_spatial <- ggplot(pred_grid_andv, aes(x = x, y = y, fill = ANDV_pred)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted\nANDV PD", option = "viridis") +
  geom_point(data = data_andv, aes(x = x, y = y, fill = NULL), 
             color = "white", size = 1, alpha = 0.5) +
  labs(title = "GAM Spatial Effect: ANDV Phylogenetic Richness",
       subtitle = "s(x, y) smooth term visualization",
       x = "Longitude", y = "Latitude") +
  theme_bw() +
  coord_fixed()

print(p_andv_spatial)
ggsave("GAM_Spatial_Effect_ANDV.png", p_andv_spatial, width = 8, height = 10, dpi = 300)
```

### 9.8 Compare OLS vs GAM Coefficient Estimates
```{r compare_ols_gam}

# Extract OLS coefficients
ols_sr_coef <- summary(ols_sr)$coefficients
ols_andv_coef <- summary(ols_andv)$coefficients

# Extract GAM coefficients
gam_sr_coef <- summary(gam_sr)$p.table
gam_andv_coef <- summary(gam_andv)$p.table

# SR Model comparison
predictors_sr <- c("PD", "suit", "Temp", "Prec", "distcent", "dist_to_centroid", "rod")

sr_comparison <- data.frame(
  Predictor = predictors_sr,
  OLS_Est = round(ols_sr_coef[predictors_sr, "Estimate"], 4),
  OLS_P = round(ols_sr_coef[predictors_sr, "Pr(>|t|)"], 4),
  GAM_Est = round(gam_sr_coef[predictors_sr, "Estimate"], 4),
  GAM_P = round(gam_sr_coef[predictors_sr, "Pr(>|z|)"], 4)
)

sr_comparison$OLS_Sig <- ifelse(sr_comparison$OLS_P < 0.05, "*", "")
sr_comparison$GAM_Sig <- ifelse(sr_comparison$GAM_P < 0.05, "*", "")

print(sr_comparison)

# ANDV Model comparison
predictors_andv <- c("PD", "suit", "Temp", "Prec", "distcent", "dist_to_centroid")

andv_comparison <- data.frame(
  Predictor = predictors_andv,
  OLS_Est = round(ols_andv_coef[predictors_andv, "Estimate"], 4),
  OLS_P = round(ols_andv_coef[predictors_andv, "Pr(>|t|)"], 4),
  GAM_Est = round(gam_andv_coef[predictors_andv, "Estimate"], 4),
  GAM_P = round(gam_andv_coef[predictors_andv, "Pr(>|t|)"], 4)
)

andv_comparison$OLS_Sig <- ifelse(andv_comparison$OLS_P < 0.05, "*", "")
andv_comparison$GAM_Sig <- ifelse(andv_comparison$GAM_P < 0.05, "*", "")

print(andv_comparison)

# Save comparisons
write.csv(sr_comparison, "SR_OLS_vs_GAM_Comparison.csv", row.names = FALSE)
write.csv(andv_comparison, "ANDV_OLS_vs_GAM_Comparison.csv", row.names = FALSE)
```



